---
title: <center>Community-AMARETTO `r ComTitle` Report</center>
pagetitle: Community-AMARETTO `r ComTitle` Report
output: 
  html_document:
    theme: cerulean
    highlight: NULL
---

<style>
  .main-container {
    max-width:  100% !important;
    padding-right: 1% !important;
    padding-left: 1% !important;
    margin-right: 1% !important;
    margin-left: 1% !important;
    table-layout: fixed !important;
  }
</style>

<center>

## Community Network Visualisation
<br><br><br>
```{r network,echo=FALSE}
cAMARETTO_ColorOneModule(cAMARETTOnetworkM,cAMARETTOnetworkC,ComNr)
```
<br><br><br>

## Regulatory Modules in Community
<br><br><br>
```{r Modules,echo=FALSE}
DTML
```
<br><br><br>

## Driver and Target Genes in Community
<br><br><br>
```{r GeneModule,echo=FALSE}
DTGenes
```
<br><br><br>

## Enrichments of Functional Categories in Community
<br><br><br>
```{r CommunityGSEA,echo=FALSE}
DTGSEA
```
<br><br><br>

## Enrichments of Driver Perturbations in Community
<br><br><br>
```{r CommunityGSEAgp,echo=FALSE}
DTGSEA_gp
```
<br><br><br>

## Enrichments of Chemical Perturbations in Community
<br><br><br>
```{r CommunityGSEAcp,echo=FALSE}
DTGSEA_cp
```
<br><br><br>

## Association of Phenotypes to Community
<br><br><br>
```{r CommunityDTPh,echo=FALSE}
DTPhC
```
<br><br><br>

</center>